---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- Knit with: devtools::build_readme() or rmarkdown::render("README.Rmd") -->
<!-- Full knit takes ~10 min (Stan MCMC + censoring test). Output is cached. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%",
  dpi       = 150
)
```

# rgetne

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://opensource.org/licenses/MIT)
[![Shiny App](https://img.shields.io/badge/Shiny-online-orange)](https://southnecrgetne.shinyapps.io/rgetne/)
<!-- badges: end -->

**Bayesian Promotion Time Cure Model -- Toolkit and Interactive App**

`rgetne` implements **Bayesian promotion-time cure models** (Yakovlev & Tsodikov, bounded cumulative hazard formulation) via [Stan](https://mc-stan.org/). It was developed as the reproducibility companion for the paper *Perioperative Chemotherapy and Long-Term Cure in Resected Grade 3 Gastroenteropancreatic Neuroendocrine Carcinomas: The GETNE/SOUTH-NEC Study* by the GETNE (Spanish Group of Neuroendocrine and Endocrine Tumors).

*(Manuscript under review -- citation and DOI will be added upon publication.)*

An interactive version of the app is available at <https://southnecrgetne.shinyapps.io/rgetne/>.

---

## Why a cure model?

In localised cancer after curative-intent surgery, conventional Cox regression tells us *whether* a treatment improves survival but not *how*. The hazard ratio cannot distinguish between:

- **Delaying recurrence** in patients who will eventually relapse (a *kinetic* effect), and
- **Eradicating residual disease** so that a subset of patients is permanently cured (a *clonogenic* effect).

The **promotion-time cure model** separates these two mechanisms. It models the unobservable number of residual tumour cell clusters (clonogenic foci) and the time each takes to produce a detectable recurrence:

$$S(t \mid \mathbf{x}, \mathbf{z}) = \exp\bigl\{-\theta(\mathbf{x})\,\bigl[1 - S_0(t \mid \mathbf{z})\bigr]\bigr\}$$

| Component | Meaning | What covariates tell us |
|---|---|---|
| **Clonogenic** $\theta(\mathbf{x})$ | Expected residual tumour burden | Which factors increase or decrease the probability of being *cured* |
| **Kinetic** $S_0(t \mid \mathbf{z})$ | Weibull latency (speed of recurrence) | Which factors accelerate or delay relapse among *uncured* patients |

The cure probability is simply $P(\text{cure}) = e^{-\theta}$: the fewer residual foci, the higher the chance that none remains viable.

This package provides:

- Full Bayesian estimation via HMC-NUTS (Stan)
- Support for **interactions** (e.g. Ki-67 x treatment) and **per-variable priors on a common, standardised scale** (see [Per-variable priors](#per-variable-priors-and-predictor-scaling))
- **Marginal predictive effects** via `posterior_epred_cure()` on arbitrary clinical-scale `newdata`
- **Kaplan-Meier** plots with optional stratification
- **Descriptive data summaries** and variable distributions
- **Identifiability diagnostics**: posterior cross-correlation, incremental censoring sensitivity
- **Clinical plots**: cure probability vs continuous covariates, cure odds ratios, population survival curves
- **Leave-one-out cross-validation** (LOO-IC)
- An **interactive Shiny application** for point-and-click exploration

---

## Installation

```r
# install.packages("remotes")
remotes::install_github("albertocarm/rgetne")
```

**Requirements:** R >= 4.1, a working C++ toolchain (for Stan), and the dependencies declared in DESCRIPTION (`rstan` >= 2.26, `shiny`, `bslib`, `survival`, `loo`, `bayesplot`, `ggplot2`, `gridExtra`).

---

## Quick start: interactive app

```r
library(rgetne)
cure()
```

This launches the Shiny application in your browser. The app auto-loads a bundled example dataset (stage I--III GEP-NEC from the GETNE registry), lets you select covariates, add interactions, run the Stan sampler, and browse all results.

---

## Full worked example

### 1. Load and preprocess data

```{r load-data}
library(rgetne)

dat <- prep_data(readRDS(example_data()))
cat(sprintf("n = %d patients, %d events (%.0f%%)\n",
            nrow(dat), sum(dat$.status_bin),
            100 * mean(dat$.status_bin)))
```

### 2. Descriptive statistics

```{r descriptives}
data_summary_table(dat)
```

### 3. Ki-67 distribution

```{r hist-ki67, fig.height=4, fig.width=6}
hist(dat$ki67_percent, breaks = 20,
     col = "#4A90D9", border = "white",
     main = "Ki-67 proliferation index",
     xlab = "Ki-67 (%)", ylab = "Frequency")
```

### 4. Kaplan-Meier curves

```{r km-overall, fig.height=5, fig.width=7}
plot_km(dat, title = "Event-free survival (overall)")
```

```{r km-treatment, fig.height=5, fig.width=7}
plot_km(dat,
        strata = "periop_therapy",
        title  = "Event-free survival by perioperative therapy")
```

```{r km-stage, fig.height=5, fig.width=7}
plot_km(dat,
        strata = "stage_III",
        title  = "Event-free survival by stage III")
```

`plot_km()` also supports interaction exploration out of the box on the
publication dataset (`publication_data()`), which ships the factor-coded
variables used throughout the manuscript. Stratify by one covariate, facet
by a second and restrict the analysis to a subset in a single call:

```{r km-facet, eval=FALSE}
pub <- readRDS(publication_data())

# One panel per tumour site, coloured by perioperative chemotherapy.
plot_km(pub,
        time_col   = "EFS_Time", status_col = "EFS_Status",
        strata     = "Perioperative_Chemo",
        facet_by   = "Site_4_Groups",
        title      = "EFS by site and perioperative chemotherapy")

# Same but restricted to colon and pancreas, split by TNM stage.
plot_km(pub,
        time_col      = "EFS_Time", status_col = "EFS_Status",
        strata        = "TNM_Stage",
        facet_by      = "Site_4_Groups",
        filter_var    = "Site_4_Groups",
        filter_values = c("Colon", "Pancreas"))
```

### 4b. Multivariable Cox models with forest plots and a combined gt table

The `cox_*` helpers fit a multivariable `rms::cph()` model and build a
publication-ready forest plot and a side-by-side gt table for both endpoints
(Event-Free and Overall Survival):

```{r cox-dual, eval=FALSE}
pub  <- readRDS(publication_data())
vars <- c("Sex", "TNM_Stage", "Primary_Surgery", "Site_4_Groups",
          "Ki67_Index", "Perioperative_Chemo", "ECOG_PS")

res <- cox_dual_endpoint(pub, vars)
res$plot_efs   # ggplot forest plot, EFS
res$plot_os    # ggplot forest plot, OS
res$table      # gt combined HR (95% CI) + p-value table
```

The same workflow is exposed in the Shiny app under the **Cox models** tab
(covariate multi-select, forest plots and combined gt table) and the
redesigned **Kaplan-Meier** tab (outcome selector, stratification, faceting,
subset filter and log-rank p-values per panel).

### 5. Define and fit the cure model

```{r fit-model, message=FALSE, results='hide'}
mod <- cure_model(
  dat,
  clonogenic = ~ primary_surgery_yes + stage_II + stage_III +
                 site_colorectal + ki67_percent * periop_therapy,
  kinetic    = ~ ki67_percent + periop_therapy,
  prior_clonogenic = 2.5,
  prior_kinetic    = 0.25,
  prior_intercept  = 2.5
)

cfit <- fit_model(mod, chains = 4, iter = 4000, warmup = 1000,
                  adapt_delta = 0.95)
```

### 6. MCMC diagnostics

```{r mcmc-checks}
mcmc_checks(cfit)
```

### 7. Coefficient summary

```{r summary-coefs}
summary_coefs(cfit)
```

### 8. Leave-one-out cross-validation

```{r loo}
loo_fit(cfit)
```

### 9. Identifiability diagnostics

```{r correlation}
test_correlation(cfit)
```

```{r censoring, message=FALSE, results='hide'}
test_censoring(cfit, chains = 2, iter = 1000)
```

### 10. Contrasts (interactions supported)

```{r contrasts}
cure_contrast(cfit, "clonogenic",
              var    = "periop_therapy",
              levels = list(periop_therapy = 1),
              ref    = list(periop_therapy = 0),
              at     = list(ki67_percent = c(30, 50, 70, 90)))
```

### 11. Clinical plots

```{r plot-cure-ref, fig.height=4, fig.width=10}
plot_continuous_effect(cfit, x_var = "ki67_percent",
                       fix = list(periop_therapy = 0),
                       xlab = "Ki-67 (%)")
```

```{r plot-cure-trt, fig.height=4, fig.width=10}
plot_continuous_effect(cfit, x_var = "ki67_percent",
                       fix = list(periop_therapy = 1),
                       xlab = "Ki-67 (%)")
```

```{r plot-or, fig.height=5, fig.width=12}
plot_contrast_cure_or(
  cfit,
  var    = "periop_therapy",
  levels = list(periop_therapy = 1),
  ref    = list(periop_therapy = 0),
  at_var = "ki67_percent",
  at_seq = seq(25, 95, by = 5),
  xlab   = "Ki-67 (%)",
  title  = "Cure OR: perioperative therapy vs none")
```

```{r plot-surv, fig.height=5, fig.width=7}
plot_surv_curve(cfit)
```

---

## Data requirements

`prep_data()` transforms raw registry data into the model-ready format:

| Raw column | Transformed | Description |
|---|---|---|
| `time_efs` | `EFS_time` | Event-free survival time (years) |
| `event_status` | `.status_bin` | 1 = event, 0 = censored |
| `primary_surgery` (Yes/No) | `primary_surgery_yes` | Primary surgery performed |
| `disease_stage` (I/II/III) | `stage_II`, `stage_III` | Disease stage dummies (ref: stage I) |
| `primary_site` (Colorectal/Other/Pancreas) | `site_colorectal`, `site_pancreas` | Primary site dummies (ref: Other) |
| `ki67_percent` | `ki67_percent` | Ki-67 proliferation index (0--100) |
| `periop_therapy` (0/1) | `periop_therapy` | Received perioperative chemotherapy |

You can upload your own `.rds` file through the Shiny app's Data panel, provided it contains these columns (or their pre-transformed equivalents).

---

## Per-variable priors and predictor scaling

When predictors live on wildly different scales (e.g. Ki-67 in 0--100 % next to a 0/1 binary) a single shared prior SD cannot be informative for all of them. The legacy entry point `cure_model()` standardises the design matrix internally before sampling, which means a prior SD of 1 is applied to the scaled coefficients and its practical meaning depends on the SD of every covariate. The newer pipeline makes scaling and priors explicit and per-variable.

### The preprocessing helpers

`escalar_continua(x, metodo)` scales a single vector:

* continuous inputs are centred and divided by `2 * SD` (`metodo = "gelman"`, default) or by one `SD` (`metodo = "zscore"`),
* binary and categorical inputs are returned unchanged,
* the original mean and scale divisor are attached as attributes so the transformation can be reproduced on new data.

`preprocess_predictors(data, predictors, metodo)` applies `escalar_continua()` to every column listed in `predictors` (the response is never touched) and returns the transformed `data.frame` plus the `center` / `scale` vectors of every continuous predictor.

```{r preprocess-example, eval=FALSE}
pp <- preprocess_predictors(dat,
                            predictors = c("ki67_percent", "stage_II",
                                           "stage_III", "site_colorectal",
                                           "periop_therapy"),
                            metodo     = "gelman")

head(pp$data[, c("ki67_percent", "stage_II", "periop_therapy")])  # scaled / 0-1
pp$center["ki67_percent"]   # original mean
pp$scale ["ki67_percent"]   # 2 * SD
```

### Fitting with `fit_cure_bayes()`

`fit_cure_bayes()` wires preprocessing, per-variable priors and Stan together. Priors can be a scalar (shared across every coefficient), a positional numeric vector, or a named list, where each name is grep-matched against the design-matrix columns (main effects *and* interactions):

```{r fit-bayes-priors, eval=FALSE}
cfit <- fit_cure_bayes(
  data       = dat,
  clonogenic = ~ primary_surgery_yes + stage_II + stage_III +
                 site_colorectal + ki67_percent * periop_therapy,
  kinetic    = ~ ki67_percent + periop_therapy,
  metodo     = "gelman",

  # One prior SD per variable, on the STANDARDISED scale.
  # Continuous (Gelman-scaled) coefficients at N(0, 1);
  # binary/categorical at N(0, 2.5); tight ridge on the interaction.
  prior_clonogenic = list(
    ki67_percent                    = 1.0,
    primary_surgery_yes             = 2.5,
    stage_II                        = 2.5,
    stage_III                       = 2.5,
    site_colorectal                 = 2.5,
    periop_therapy                  = 2.5,
    "ki67_percent:periop_therapy"   = 0.5   # regularise the interaction
  ),
  prior_kinetic = list(
    ki67_percent   = 0.5,
    periop_therapy = 1.0
  ),
  prior_intercept = 2.5,

  chains = 4, iter = 4000, warmup = 1000, adapt_delta = 0.95
)
```

When `prior_clonogenic` or `prior_kinetic` is left at its default (`NULL`), `fit_cure_bayes()` applies an asymmetric default aligned with the biology of the model. The **clonogenic (cure) part** keeps a shared weakly-informative `N(0, 2.5)` for every coefficient, so the cure-probability regression stays unconstrained. The **kinetic (latency) part** instead defaults to `N(0, 0.25)` for `ki67_percent` and `periop_therapy` (including their interactions), `N(0, 1)` for other standardised continuous coefficients, and `N(0, 2.5)` for the remaining binary / categorical ones — reflecting the prior belief that proliferation and perioperative therapy should only modulate the hazard shape moderately once a subject is at risk of recurrence.

### Marginal predictive effects

The fitted object stores the scaling metadata at `cfit$spec$scaling`, which lets `posterior_epred_cure()` project an arbitrary `newdata` on the **original clinical scale** into the same standardised space, compute the posterior linear predictor, and return one of the four quantities `"cure_prob"`, `"theta"`, `"lambda"`, `"median_time"` as a `n_draws x n_rows` matrix.

```{r epred-example, eval=FALSE}
newdata <- expand.grid(
  ki67_percent        = seq(5, 90, by = 5),   # clinical scale
  periop_therapy      = c(0, 1),
  primary_surgery_yes = 1,
  stage_II            = 0,
  stage_III           = 0,
  site_colorectal     = 0
)

cure_draws <- posterior_epred_cure(cfit, newdata, type = "cure_prob")
summary_tbl <- summarize_epred(cure_draws)    # median + 95% CrI per row
head(summary_tbl)
```

The returned `summary_tbl` carries back the original-scale `newdata` so the estimates can be plotted directly against clinical values (e.g. Ki-67 in %) without manual back-transformation.

### Interactive counterpart

The same inputs are exposed in the Shiny app under the sidebar panels `"Clonogenic Model"` and `"Kinetic Model"` (one numeric input per selected variable) and in the new tab `"Marginal effects"`, which wraps `posterior_epred_cure()` + `summarize_epred()` with a 95% credible ribbon.

---

## Package functions

| Function | Description |
|---|---|
| `cure()` / `cure_app()` | Launch the interactive Shiny application |
| `prep_data()` | Preprocess raw data for modelling |
| `escalar_continua()` | Scale a continuous vector (Gelman or Z-score); leaves binary / categorical inputs untouched |
| `preprocess_predictors()` | Apply `escalar_continua()` to predictor columns only and record the original mean / scale of every continuous predictor |
| `cure_model()` | Define a model specification (formulas + priors); `scale = FALSE` skips the internal standardisation |
| `fit_model()` | Fit a specification via Stan HMC-NUTS |
| `fit_cure_bayes()` | End-to-end wrapper: preprocess predictors, build weakly-informative per-variable priors, fit Stan and attach the scaling metadata |
| `posterior_epred_cure()` | Posterior expected prediction on clinical-scale `newdata` (`cure_prob`, `theta`, `lambda`, `median_time`) |
| `summarize_epred()` | Collapse a draws x rows prediction matrix into median, mean and credible interval |
| `mcmc_checks()` | MCMC convergence diagnostics (R-hat, ESS, divergences) |
| `summary_coefs()` | Posterior coefficient summary on the original scale |
| `loo_fit()` | Leave-one-out cross-validation |
| `cure_contrast()` | Compute scenario contrasts with credible intervals |
| `test_correlation()` | Posterior cross-correlation between sub-models |
| `test_censoring()` | Incremental censoring sensitivity test |
| `plot_km()` | Kaplan-Meier survival curves (stratify, facet, subset, log-rank p) |
| `cox_multivar_fit()` / `cox_forest_plot()` / `cox_combined_table()` | Multivariable Cox model, forest plot and publication gt table |
| `cox_dual_endpoint()` | One-shot wrapper: fit EFS + OS, build both forests and the combined table |
| `publication_data()` | Path to the factor-coded publication dataset used by the KM and Cox tabs |
| `plot_continuous_effect()` | Triptych: clonogenic load, cure rate, median failure time |
| `plot_contrast_cure_or()` | Cure odds ratio across a continuous modifier |
| `plot_surv_curve()` | Population survival curve from the cure model |
| `data_summary_table()` | Descriptive statistics table |
| `example_data()` / `stan_model_file()` | Paths to bundled assets |

---

## Citation

> *Manuscript under review.* Full citation and DOI will be added upon publication.

> rgetne: Bayesian Promotion Time Cure Model Toolkit (R package). https://github.com/albertocarm/rgetne

---

## License

MIT. See [LICENSE](LICENSE).
