---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- Knit with: devtools::build_readme() or rmarkdown::render("README.Rmd") -->
<!-- Full knit takes ~10 min (Stan MCMC + censoring test). Output is cached. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%",
  dpi       = 150
)
```

# rgetne

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://opensource.org/licenses/MIT)
[![Shiny App](https://img.shields.io/badge/Shiny-online-orange)](https://southnecrgetne.shinyapps.io/rgetne/)
<!-- badges: end -->

**Bayesian Promotion Time Cure Model -- Toolkit and Interactive App**

`rgetne` implements **Bayesian promotion-time cure models** (Yakovlev & Tsodikov, bounded cumulative hazard formulation) via [Stan](https://mc-stan.org/). It was developed as the reproducibility companion for the paper *Perioperative Chemotherapy and Long-Term Cure in Resected Grade 3 Gastroenteropancreatic Neuroendocrine Carcinomas: The GETNE/SOUTH-NEC Study* by the GETNE (Spanish Group of Neuroendocrine and Endocrine Tumors).

*(Manuscript under review -- citation and DOI will be added upon publication.)*

An interactive version of the app is available at <https://southnecrgetne.shinyapps.io/rgetne/>.

---

## Why a cure model?

In localised cancer after curative-intent surgery, conventional Cox regression tells us *whether* a treatment improves survival but not *how*. The hazard ratio cannot distinguish between:

- **Delaying recurrence** in patients who will eventually relapse (a *kinetic* effect), and
- **Eradicating residual disease** so that a subset of patients is permanently cured (a *clonogenic* effect).

The **promotion-time cure model** separates these two mechanisms. It models the unobservable number of residual tumour cell clusters (clonogenic foci) and the time each takes to produce a detectable recurrence:

$$S(t \mid \mathbf{x}, \mathbf{z}) = \exp\bigl\{-\theta(\mathbf{x})\,\bigl[1 - S_0(t \mid \mathbf{z})\bigr]\bigr\}$$

| Component | Meaning | What covariates tell us |
|---|---|---|
| **Clonogenic** $\theta(\mathbf{x})$ | Expected residual tumour burden | Which factors increase or decrease the probability of being *cured* |
| **Kinetic** $S_0(t \mid \mathbf{z})$ | Weibull latency (speed of recurrence) | Which factors accelerate or delay relapse among *uncured* patients |

The cure probability is simply $P(\text{cure}) = e^{-\theta}$: the fewer residual foci, the higher the chance that none remains viable.

This package provides:

- Full Bayesian estimation via HMC-NUTS (Stan)
- Support for **interactions** (e.g. Ki-67 x treatment) and **per-variable priors**
- **Kaplan-Meier** plots with optional stratification
- **Descriptive data summaries** and variable distributions
- **Identifiability diagnostics**: posterior cross-correlation, incremental censoring sensitivity
- **Clinical plots**: cure probability vs continuous covariates, cure odds ratios, population survival curves
- **Leave-one-out cross-validation** (LOO-IC)
- An **interactive Shiny application** for point-and-click exploration

---

## Installation

```r
# install.packages("remotes")
remotes::install_github("albertocarm/rgetne")
```

**Requirements:** R >= 4.1, a working C++ toolchain (for Stan), and the dependencies declared in DESCRIPTION (`rstan` >= 2.26, `shiny`, `bslib`, `survival`, `loo`, `bayesplot`, `ggplot2`, `gridExtra`).

---

## Quick start: interactive app

```r
library(rgetne)
cure()
```

This launches the Shiny application in your browser. The app auto-loads a bundled example dataset (stage I--III GEP-NEC from the GETNE registry), lets you select covariates, add interactions, run the Stan sampler, and browse all results.

---

## Full worked example

### 1. Load and preprocess data

```{r load-data}
library(rgetne)

dat <- prep_data(readRDS(example_data()))
cat(sprintf("n = %d patients, %d events (%.0f%%)\n",
            nrow(dat), sum(dat$.status_bin),
            100 * mean(dat$.status_bin)))
```

### 2. Descriptive statistics

```{r descriptives}
data_summary_table(dat)
```

### 3. Ki-67 distribution

```{r hist-ki67, fig.height=4, fig.width=6}
hist(dat$ki67_percent, breaks = 20,
     col = "#4A90D9", border = "white",
     main = "Ki-67 proliferation index",
     xlab = "Ki-67 (%)", ylab = "Frequency")
```

### 4. Kaplan-Meier curves

`plot_km()` draws shaded 95% confidence bands, censoring tick marks, an
at-risk table beneath the curves and a log-rank `p`-value when stratified.
Stratification by binary, factor or character columns with any number of
levels is supported.

```{r km-overall, fig.height=5.5, fig.width=7}
plot_km(dat, title = "Event-free survival (overall)")
```

```{r km-treatment, fig.height=5.5, fig.width=7}
plot_km(dat,
        strata        = "periop_therapy",
        strata_labels = c("periop_therapy = 0" = "No periop. chemotherapy",
                          "periop_therapy = 1" = "Periop. chemotherapy"),
        title         = "Event-free survival by perioperative therapy")
```

```{r km-stage, fig.height=5.5, fig.width=7}
plot_km(dat,
        strata        = "stage_III",
        strata_labels = c("stage_III = 0" = "Stage I-II",
                          "stage_III = 1" = "Stage III"),
        title         = "Event-free survival by stage III")
```

```{r km-site, fig.height=6, fig.width=7.5}
# Stratify by tumour location (factor, multiple levels). Uses the original
# primary_site factor before it is collapsed into 0/1 dummies by prep_data().
raw <- readRDS(example_data())
dat_site <- prep_data(raw)
dat_site$primary_site <- factor(raw$primary_site[match(rownames(dat_site),
                                                       rownames(raw))])
plot_km(dat_site,
        strata = "primary_site",
        title  = "Event-free survival by primary tumour location")
```

### 5. Define and fit the cure model

```{r fit-model, message=FALSE, results='hide'}
mod <- cure_model(
  dat,
  clonogenic = ~ primary_surgery_yes + stage_II + stage_III +
                 site_colorectal + ki67_percent * periop_therapy,
  kinetic    = ~ ki67_percent + periop_therapy,
  prior_clonogenic = 2.5,
  prior_kinetic    = 0.25,
  prior_intercept  = 2.5
)

cfit <- fit_model(mod, chains = 4, iter = 4000, warmup = 1000,
                  adapt_delta = 0.95)
```

### 6. MCMC diagnostics

```{r mcmc-checks}
mcmc_checks(cfit)
```

### 7. Coefficient summary

```{r summary-coefs}
summary_coefs(cfit)
```

### 8. Leave-one-out cross-validation

```{r loo}
loo_fit(cfit)
```

### 9. Identifiability diagnostics

```{r correlation}
test_correlation(cfit)
```

```{r censoring, message=FALSE, results='hide'}
test_censoring(cfit, chains = 2, iter = 1000)
```

### 10. Contrasts (interactions supported)

```{r contrasts}
cure_contrast(cfit, "clonogenic",
              var    = "periop_therapy",
              levels = list(periop_therapy = 1),
              ref    = list(periop_therapy = 0),
              at     = list(ki67_percent = c(30, 50, 70, 90)))
```

### 11. Clinical plots

```{r plot-cure-trt, fig.height=4, fig.width=10}
plot_continuous_effect(cfit, x_var = "ki67_percent",
                       fix  = list(periop_therapy = 1),
                       xlab = "Ki-67 (%)")
```

```{r plot-or, fig.height=5, fig.width=12}
plot_contrast_cure_or(
  cfit,
  var    = "periop_therapy",
  levels = list(periop_therapy = 1),
  ref    = list(periop_therapy = 0),
  at_var = "ki67_percent",
  at_seq = seq(25, 95, by = 5),
  xlab   = "Ki-67 (%)",
  title  = "Cure OR: perioperative therapy vs none")
```

```{r plot-surv, fig.height=5, fig.width=7}
plot_surv_curve(cfit)
```

### 12. Pairwise interaction screen (Supplementary Table A5)

`screen_interactions()` fits the additive base model and one model per
supplied pair (base + a single interaction term), then reports the
posterior median and 95% credible interval of the interaction coefficient
on the original (un-standardised) covariate scale, the probability of
direction, and the LOO-CV ELPD difference relative to the additive base
model. Results are sorted by strength of evidence.

```{r interaction-screen, eval=FALSE}
base_clono <- ~ ki67_percent + periop_therapy + primary_surgery_yes +
                 stage_III   + site_pancreas  + site_colorectal

a5 <- screen_interactions(
  data            = dat,
  base_clonogenic = base_clono,
  kinetic         = ~ ki67_percent + periop_therapy,
  pairs = list(
    c("ki67_percent",        "periop_therapy"),
    c("ki67_percent",        "primary_surgery_yes"),
    c("ki67_percent",        "site_pancreas"),
    c("periop_therapy",      "site_pancreas"),
    c("primary_surgery_yes", "stage_III"),
    c("primary_surgery_yes", "site_colorectal"),
    c("periop_therapy",      "primary_surgery_yes"),
    c("periop_therapy",      "stage_III"),
    c("stage_III",           "site_pancreas"),
    c("primary_surgery_yes", "site_pancreas")
  ),
  labels = c(
    "ki67_percent:periop_therapy"         = "Ki-67 \u00d7 Periop. chemotherapy",
    "ki67_percent:primary_surgery_yes"    = "Ki-67 \u00d7 Primary surgery",
    "ki67_percent:site_pancreas"          = "Ki-67 \u00d7 Pancreatic site",
    "periop_therapy:site_pancreas"        = "Periop. chemo \u00d7 Pancreatic site",
    "primary_surgery_yes:stage_III"       = "Primary surgery \u00d7 Stage III",
    "primary_surgery_yes:site_colorectal" = "Primary surgery \u00d7 Other site",
    "periop_therapy:primary_surgery_yes"  = "Periop. chemo \u00d7 Primary surgery",
    "periop_therapy:stage_III"            = "Periop. chemo \u00d7 Stage III",
    "stage_III:site_pancreas"             = "Stage III \u00d7 Pancreatic site",
    "primary_surgery_yes:site_pancreas"   = "Primary surgery \u00d7 Pancreatic site"
  ),
  chains = 4, iter = 4000, warmup = 1000
)

knitr::kable(format_interaction_table(a5), align = "lrlrrl",
             caption = "Supplementary Table A5. Pairwise interaction screen.")
```

The full screen yields the table below. Each interaction was fitted as a
separate model and compared against the additive main-effects model via
LOO-CV.

| Interaction term                       | Coefficient | 95% CrI         | P(direction) | ΔELPD | Evidence       |
|----------------------------------------|------------:|:----------------|-------------:|------:|:---------------|
| Ki-67 × Periop. chemotherapy           |      −0.014 | −0.031, 0.005   |        93.5% |  −0.4 | Suggestive **  |
| Ki-67 × Primary surgery                |      −0.011 | −0.032, 0.009   |        87.4% |  −0.5 | Suggestive *   |
| Ki-67 × Pancreatic site                |       0.010 | −0.011, 0.032   |        84.2% |  −1.5 | Suggestive *   |
| Periop. chemo × Pancreatic site        |       0.576 | −0.40, 1.61     |        86.8% |  −0.7 | Suggestive *   |
| Primary surgery × Stage III            |      −1.772 | −5.24, 0.81     |        90.0% |  −0.4 | Suggestive **  |
| Primary surgery × Other site           |      −1.051 | −2.92, 0.40     |        91.8% |  −1.6 | Suggestive **  |
| Periop. chemo × Primary surgery        |      −0.383 | −1.40, 0.80     |        74.1% |  −0.8 | Weak           |
| Periop. chemo × Stage III              |       0.016 | −1.57, 1.90     |        51.0% |  −2.2 | None           |
| Stage III × Pancreatic site            |       0.158 | −1.60, 2.22     |        56.4% |  −2.6 | None           |
| Primary surgery × Pancreatic site      |      −0.456 | −2.36, 1.06     |        70.6% |  −1.6 | Weak           |

Interaction coefficients are reported on the original (unscaled)
covariate scale. For continuous × binary interactions (e.g.,
Ki-67 × chemotherapy) the coefficient represents the change in the slope
of the continuous variable when the binary variable is switched on; for
binary × binary interactions it represents the departure from additivity
on the log-θ scale. `P(direction)` is the posterior probability that the
interaction coefficient has the sign of its median (`** pd > 90%`,
`* pd > 80%`). ΔELPD is the difference in expected log pointwise
predictive density relative to the additive base model: negative values
favour the interaction model and a difference exceeding twice its
standard error is considered meaningful.

**Comment on the interactions.** No interaction model meaningfully
improved predictive performance over the additive specification —
every ΔELPD is negative (favouring the simpler additive model) and well
within `2 × SE`. The strongest signal, *Ki-67 × peri-operative
chemotherapy* (`pd = 93.5%`), is a negative interaction on the
log-θ scale: the slope of Ki-67 in the clonogenic linear predictor
becomes more negative when chemotherapy is given, i.e. the cure benefit
of chemotherapy grows with proliferation index. Conditional contrasts
showed that chemotherapy had no detectable effect at Ki-67 ≤ 30%
(clonogenic ratio 0.98; CrI 0.40–2.13) but progressively reduced
clonogenic burden at higher values (Ki-67 = 70%: ratio 0.56,
CrI 0.33–0.92; Ki-67 = 90%: ratio 0.43, CrI 0.23–0.79), corresponding
to cure odds ratios of 4.2 and 10.6. *Primary surgery × Other site* and
*Primary surgery × Stage III* sit at `pd ≈ 90%` but with very wide
intervals straddling zero, so they should be read as exploratory hints
that the surgical effect is not perfectly homogeneous across sites and
stages rather than as confirmed effect modifiers. The remaining seven
pairs show no meaningful evidence of departure from additivity.

CrI = credible interval; ELPD = expected log pointwise predictive
density; LOO-CV = leave-one-out cross-validation; pd = probability of
direction; θ = expected clonogenic burden.

---

## Data requirements

`prep_data()` transforms raw registry data into the model-ready format:

| Raw column | Transformed | Description |
|---|---|---|
| `time_efs` | `EFS_time` | Event-free survival time (years) |
| `event_status` | `.status_bin` | 1 = event, 0 = censored |
| `primary_surgery` (Yes/No) | `primary_surgery_yes` | Primary surgery performed |
| `disease_stage` (I/II/III) | `stage_II`, `stage_III` | Disease stage dummies (ref: stage I) |
| `primary_site` (Colorectal/Other/Pancreas) | `site_colorectal`, `site_pancreas` | Primary site dummies (ref: Other) |
| `ki67_percent` | `ki67_percent` | Ki-67 proliferation index (0--100) |
| `periop_therapy` (0/1) | `periop_therapy` | Received perioperative chemotherapy |

You can upload your own `.rds` file through the Shiny app's Data panel, provided it contains these columns (or their pre-transformed equivalents).

---

## Package functions

| Function | Description |
|---|---|
| `cure()` / `cure_app()` | Launch the interactive Shiny application |
| `prep_data()` | Preprocess raw data for modelling |
| `cure_model()` | Define model specification (formulas + priors) |
| `fit_model()` | Fit the model via Stan HMC-NUTS |
| `mcmc_checks()` | MCMC convergence diagnostics (R-hat, ESS, divergences) |
| `summary_coefs()` | Posterior coefficient summary on the original scale |
| `loo_fit()` | Leave-one-out cross-validation |
| `cure_contrast()` | Compute scenario contrasts with credible intervals |
| `screen_interactions()` / `format_interaction_table()` | Pairwise interaction screen + Supplementary Table A5 |
| `test_correlation()` | Posterior cross-correlation between sub-models |
| `test_censoring()` | Incremental censoring sensitivity test |
| `plot_km()` | Kaplan-Meier survival curves (multi-level strata, CI bands, at-risk table, log-rank `p`) |
| `plot_continuous_effect()` | Triptych: clonogenic load, cure rate, median failure time |
| `plot_contrast_cure_or()` | Cure odds ratio across a continuous modifier |
| `plot_surv_curve()` | Population survival curve from the cure model |
| `data_summary_table()` | Descriptive statistics table |
| `example_data()` / `stan_model_file()` | Paths to bundled assets |

---

## Citation

> *Manuscript under review.* Full citation and DOI will be added upon publication.

> rgetne: Bayesian Promotion Time Cure Model Toolkit (R package). https://github.com/albertocarm/rgetne

---

## License

MIT. See [LICENSE](LICENSE).
